---
title: "Introduction to Support Vector Clustering with SVclust"
author: "Lucas Scheffer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Support Vector Clustering with SVclust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
  
## Theory  
  
Support Vector Clustering (SVC) is a clustering algorithm which adapts the Support Vector Machines (SVM) theory, proposed by Vapnik *et al*, to cluster the datapoints into groups with similar characteristics.  
  
Developed by Ben-Hur *et al.*(2001), this method maps data points to a high-dimensional feature space using a kernel $\phi$. In this space, we seek the smallest sphere that encloses the images of the data points. We can consider doing this task by minimizing the hypersphere radius $R$:  
  
$$
\min_{R} R^2 \text{.}
$$
  
To account for outliers or overlapping distributions, we introduce slack variables $\xi$ and a penalty constant $C$. The primal optimization problem is defined as:
  
$$
R^2 + C \sum_{i=1}^n \xi_i\text{,}
$$
  
subject to: $$\|\Phi(\mathbf{x}) - a\|^2 \le R^2 + \xi_i\text{ , }\xi_i\geq0$$, with $a$ as the center of hypersphere.  
  
Points that lie on the surface of the sphere are the **Support Vectors (SVs)**, while points outside are **Bounded Support Vectors (BSVs)**.  
  
### The Dual Problem  
  
To solve the primal optimization problem efficiently, we use the Lagrangian formulation. By introducing the Lagrange multipliers $\beta_i$ associated with the constraints, we derive the following dual problem, which is a Quadratic Programming (QP) task:  
  
$$
\max_{\beta} \sum_{i=1}^n \beta_i K(\mathbf{x}, \mathbf{x}) - \sum_{i,j=1}^n \beta_i \beta_j K(\mathbf{x}, \mathbf{x}'),
$$  
  
subject to:  
  
$$
0 \le \beta_i \le C, \quad \sum_{i=1}^n \beta_i = 1,
$$
  
where $K(\mathbf{x}, \mathbf{x}') = \Phi(\mathbf{x}) \cdot \Phi(\mathbf{x}')$ is the kernel function (e.g., Gaussian, Laplacian, or Cauchy).   
  
The solution to this problem yields the optimal multipliers $\beta_i$. The distance of any point $x$ to the center of the sphere in the feature space, $R^2(x)$, can then be calculated as:  
  
$$R^2(x) = K(x, x) - 2 \sum_{i=1}^n \beta_i K(\mathbf{x}, x) + \sum_{i,j=1}^n \beta_i \beta_j K(\mathbf{x}, \mathbf{x}')$$  
  
The radius of the sphere, $R$, is determined by the distance $R(\mathbf{x})$ for any Support Vector $\mathbf{x}$ (where $0 < \beta_i < C$).  
  
## Connection Step  
  
Two points belong to the same cluster if the path connecting them in the feature space remains within the sphere. This is checked numerically by sampling `scan_n` points along the segment. Two points $\mathbf{x}$ and $\mathbf{x}'$ are connected if $||\phi(t\mathbf{x} + (1 −t)\mathbf{x}') −a||^2 \leq R^2\text{, } \forall t \in (0,1)$, with $t$ as a position in the segment.  
  
## Parameter Selection  
  
The Gaussian kernel parameter $\gamma$ (gamma) is the most critical hyperparameter in SVC. It controls the scale at which the data is probed:  
  
-   A small $\gamma$ results in a smooth feature space, often leading to a single large cluster.  
  
-   A large $\gamma$ increases the tightness of the hypersphere, which may lead to over-fragmentation.  
  
`SVclust` provides an automated $\gamma$ estimation method based on the median of pairwise squared distances, which is a common heuristic in kernel methods. Users can also specify $\gamma$ directly for more control.
  
## Computational Efficiency  
  
Since calculating the adjacency matrix can be $O(n^2)$, this package offers a hybrid approach for scalability:  
  
-   Sampling: Run SVC on a manageable subset (defined by `n_row`).  
  
-   SVM Extension: Classify the remaining points using an SVM trained on the SVC results.  
  
## Example: non-linear clustering  
  
```{r setup}
## Sintetic dataset (two circles)
# library(SVclust)
set.seed(42)
n <- 100
theta <- seq(0, 2*pi, length.out = n)
c1_data <- cbind(cos(theta), sin(theta)) + matrix(runif(n*2, -0.1, 0.1), ncol=2)
c2_data <- cbind(2*cos(theta), 2*sin(theta)) + matrix(runif(n*2, -0.1, 0.1), ncol=2)

df_circles <- as.data.frame(rbind(c1_data, c2_data))
colnames(df_circles) <- c("c1", "c2")
```

```{r comparison, fig.width = 10, fig.height = 5, out.width = "100%"}
results <-  SVC(
  df_circles,
  gamma = 2,
  scan_n = 10,
  kernel = "gaussian", 
  verbose = FALSE,
  n_cores = 2
)
library(ggplot2)
set.seed(42)

df_circles$svc_cluster <- as.factor(results$clusters)
p_svc <- ggplot(df_circles, aes(x = c1, y = c2, color = svc_cluster)) +
  geom_point(size = 2) + coord_fixed() +
  labs(title = "Support Vector Clustering", x = "C1", y = "C2", color = "SVC Cluster") +
  theme_minimal() + theme(legend.position = "bottom")

p_svc
# library(gridExtra)
# km_res <- kmeans(df_circles[, 1:2], centers = 2)
# df_circles$kmeans_cluster <- as.factor(km_res$cluster)
# p_km <- ggplot(df_circles, aes(x = c1, y = c2, color = kmeans_cluster)) +
  # geom_point(size = 2) + coord_fixed() +
  # labs(title = "K-means Clustering", x = "C1", y = "C2", color = "KM Cluster") +
  # theme_minimal() + theme(legend.position = "bottom")
# grid.arrange(p_km, p_svc, ncol = 2)
```

## References

1.  Ben-Hur, A., Horn, D., Siegelmann, H. T., & Vapnik, V. (2001). Support vector clustering. *Journal of Machine Learning Research*, 2(Dec), 125-137.
2.  Schölkopf, B., Williamson, R. C., Smola, A. J., Shawe-Taylor, J., & Platt, J. C. (1999). Support vector method for novelty detection. *Advances in Neural Information Processing Systems*, 12.
3.  Stellato, B., Banjac, G., Goulart, P., Bemporad, A., & Boyd, S. (2020). OSQP: An Operator Splitting Solver for Quadratic Programs. *Mathematical Programming Computation*, 12(4), 637-672.
