---
title: "Introduction to Support Vector Clustering with SVclust"
author: "Lucas Scheffer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Support Vector Clustering with SVclust}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Theory

Support Vector Clustering (SVC) is a clustering algorithm which adapts the Support Vector Machines (SVM) theory, proposed by Vapnik *et al.*, to cluster the data points into groups with similar characteristics.

Developed by Ben-Hur *et al.* (2001), this method maps data points to a high-dimensional feature space using a nonlinear map $\Phi$ induced by a kernel function $K$. In this space, we seek the smallest sphere that encloses the images of the data points. We can consider doing this task by minimizing the hypersphere radius $R$:

$$
\min_{R} R^2 \text{.}
$$

To account for outliers or overlapping distributions, we introduce slack variables $\xi_i$ and a penalty constant $C$. The primal optimization problem is defined as:

$$
\min_{R, \mathbf{a}, \xi} \; R^2 + C \sum_{i=1}^n \xi_i\text{,}
$$

subject to $\|\Phi(\mathbf{x}_i) - \mathbf{a}\|^2 \le R^2 + \xi_i$ and $\xi_i \geq 0$ for $i = 1, \dots, n$, with $\mathbf{a}$ as the center of the hypersphere.

Points that lie on the surface of the sphere are the **Support Vectors (SVs)**, while points outside are **Bounded Support Vectors (BSVs)**.

### The Dual Problem

To solve the primal optimization problem efficiently, we use the Lagrangian formulation. By introducing the Lagrange multipliers $\beta_i$ associated with the constraints, we derive the following dual problem, which is a Quadratic Programming (QP) task:

$$
\max_{\beta} \sum_{i=1}^n \beta_i K(\mathbf{x}_i, \mathbf{x}_i) - \sum_{i,j=1}^n \beta_i \beta_j K(\mathbf{x}_i, \mathbf{x}_j),
$$

subject to:

$$
0 \le \beta_i \le C, \quad \sum_{i=1}^n \beta_i = 1,
$$

where $K(\mathbf{x}, \mathbf{x}') = \Phi(\mathbf{x}) \cdot \Phi(\mathbf{x}')$ is the kernel function. `SVclust` offers the Gaussian kernel $K(\mathbf{x}, \mathbf{x}') = e^{-\gamma \|\mathbf{x} - \mathbf{x}'\|^2}$, the Laplacian kernel $e^{-\gamma \|\mathbf{x} - \mathbf{x}'\|}$ and the Cauchy kernel $1 / (1 + \gamma \|\mathbf{x} - \mathbf{x}'\|^2)$.

The solution to this problem yields the optimal multipliers $\beta_i$. The distance of any point $\mathbf{x}$ to the center of the sphere in the feature space, $R^2(\mathbf{x})$, can then be calculated as:

$$R^2(\mathbf{x}) = K(\mathbf{x}, \mathbf{x}) - 2 \sum_{i=1}^n \beta_i K(\mathbf{x}_i, \mathbf{x}) + \sum_{i,j=1}^n \beta_i \beta_j K(\mathbf{x}_i, \mathbf{x}_j)$$

The radius of the sphere, $R$, is determined by the distance $R(\mathbf{x}_i)$ for any Support Vector $\mathbf{x}_i$ (where $0 < \beta_i < C$); `SVclust` uses the average of $R^2(\mathbf{x}_i)$ over the support vectors.

## Connection Step

Two points belong to the same cluster if the path connecting them in the feature space remains within the sphere. Two points $\mathbf{x}_i$ and $\mathbf{x}_j$ are connected if $\|\Phi(t\mathbf{x}_i + (1 - t)\mathbf{x}_j) - \mathbf{a}\|^2 \leq R^2$ for all $t \in [0, 1]$, with $t$ as a position in the segment. This is checked numerically at `scan_n` positions $t$ drawn uniformly at random in $[0, 1]$, the same positions for every pair (Ben-Hur *et al.* used 20 points), so call `set.seed()` before `SVC()` for reproducible results. Clusters are the connected components of the resulting adjacency graph.

## Parameter Selection

The kernel parameter $\gamma$ (gamma) is the most critical hyperparameter in SVC. It controls the scale at which the data is probed:

-   A small $\gamma$ results in a smooth feature space, often leading to a single large cluster.

-   A large $\gamma$ increases the tightness of the hypersphere, which may lead to over-fragmentation.

When `gamma = NULL` (the default) or `gamma = "auto"`, `SVclust` estimates $\gamma$ with the median heuristic: the inverse of the median pairwise squared distance (the inverse of the median pairwise distance for the Laplacian kernel). For the Gaussian kernel, this is the middle value returned by `kernlab::sigest()` on unscaled data. The heuristic places the kernel width at the overall scale of the data, where SVC usually finds a single cluster, so treat it as a starting point. Ben-Hur *et al.* (2001, Section 4.2) propose starting from a small $\gamma$ with a single cluster, such as $1 / \max_{i,j} \|\mathbf{x}_i - \mathbf{x}_j\|^2$, and increasing it to observe the formation of an increasing number of clusters, preferring solutions with a small number of support vectors that remain stable over a range of values. Users can also specify $\gamma$ directly for more control.

## Computational Efficiency

The connection step checks all $n(n-1)/2$ pairs of points and evaluates $R^2$ at `scan_n` positions on each segment, and each evaluation sums over the $n$ data points. It therefore costs $O(s\,n^3)$ kernel evaluations, for $s$ = `scan_n`, and dominates the running time; `n_cores` spreads the pairs over several R processes. For large datasets, this package offers a hybrid approach:

-   Sampling: Run SVC on a random subset of `n_obs` rows.

-   SVM Extension: Classify the remaining points using an SVM trained on the SVC results (if SVC finds a single cluster, all points are assigned to it).

## Example: non-linear clustering

```{r setup}
## Synthetic dataset (two circles)
library(SVclust)
set.seed(42)
n <- 100
theta <- seq(0, 2*pi, length.out = n)
c1_data <- cbind(cos(theta), sin(theta)) + matrix(runif(n*2, -0.1, 0.1), ncol=2)
c2_data <- cbind(2*cos(theta), 2*sin(theta)) + matrix(runif(n*2, -0.1, 0.1), ncol=2)

df_circles <- as.data.frame(rbind(c1_data, c2_data))
colnames(df_circles) <- c("c1", "c2")
```

```{r comparison, fig.width = 10, fig.height = 5, out.width = "100%"}
results <-  SVC(
  df_circles,
  gamma = 2,
  scan_n = 10,
  kernel = "gaussian",
  verbose = FALSE,
  n_cores = 2
)
table(results$clusters)

# K-means for comparison
set.seed(42)
km_res <- kmeans(df_circles, centers = 2)

library(ggplot2)
plot_df <- rbind(
  data.frame(df_circles, method = "Support Vector Clustering", cluster = factor(results$clusters)),
  data.frame(df_circles, method = "K-means", cluster = factor(km_res$cluster))
)
ggplot(plot_df, aes(x = c1, y = c2, color = cluster)) +
  geom_point(size = 2) + coord_fixed() +
  facet_wrap(~ method) +
  labs(x = "C1", y = "C2", color = "Cluster") +
  theme_minimal() + theme(legend.position = "bottom")
```

## References

1.  Ben-Hur, A., Horn, D., Siegelmann, H. T., & Vapnik, V. (2001). Support vector clustering. *Journal of Machine Learning Research*, 2(Dec), 125-137. <https://www.jmlr.org/papers/v2/horn01a.html>
2.  Schölkopf, B., Williamson, R. C., Smola, A. J., Shawe-Taylor, J., & Platt, J. C. (1999). Support vector method for novelty detection. *Advances in Neural Information Processing Systems*, 12.
3.  Stellato, B., Banjac, G., Goulart, P., Bemporad, A., & Boyd, S. (2020). OSQP: An Operator Splitting Solver for Quadratic Programs. *Mathematical Programming Computation*, 12(4), 637-672.
